\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-08 honors_thesis/journal_paper/paper.tex
tweaks in the Introduction and some rewriting in Section II
I wouldn't be shocked if you hate some of these changes, so feel free to
"cherry pick" those that you like.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) techniques commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, being careful not to make
changes that can not be undone.  If the processor guesses right, execution is
much farther ahead than if it had just idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

@@

The balance of this paper is organized as follows: \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once, then
presents a comparison of the speculative algorithm with the original PSO.  In
\sec{relax}, we discuss various methods of improving the performance of
speculative evaluation in PSO, all of which break the requirement of strictly
reproducing the behavior of the original algorithm.  In \sec{conclusion} and
\sec{future} we conclude and discuss future work.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-sis07}.  \( \ncoeff \), \( \pcoeff \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies a particle is also a
neighbor to itself in that its own position and value are considered when
updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
using the Ring topology each particle with index $i$ has three neighbors:
$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
$p$ neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in the
introduction, for some functions, and for large numbers of processors, just
adding particles reaches a point of diminishing returns.  That is, adding
processors does not help us reach any given level of fitness appreciably
faster.

To address this issue we present here a different use of additional processors.
Instead of adding particles we employ a speculative evaluation approach that
allows us to perform iterations two at a time.

To do so we must first refactor PSO such that the determination of the social
and cognitive components, $\nbest$ and $\pbest$, which depend upon the value of
the objective function, is separate from the rest of the computation.  For
simplicity in this explanation, we restrict our discussion here to the case of
a Ring topology with two neighbors which we will call the ``right neighbor''
and ``left neighbor,'' whose positions are represented as $\rightn$ and
$\leftn$ respectively.  Though we will only describe the case of the Ring
topology, our methods can easily be extended to arbitrary topologies.

Our refactoring hinges on the idea that there is a finite set of possible
updates to $\nbest$ and $\pbest$ at each iteration.  For the Ring topology we
are discussing there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update; a hyphen represents no update, $L$ represents an update
to $\nbest$ coming from the left neighbor, $R$ represents an update to $\nbest$
coming from the right neighbor, and $S$ represents an update to either $\pbest$
or $\nbest$ coming from the particle itself.  As an example, $\casexn$ refers
to the case that the particle finds a new personal best, but neither it nor its
neighbors found a position that updated its neighborhood best.  In the
equations that follow, we refer to an unspecified update case as $\casegen$,
and to the set of cases collectively as $\caseset$

@@

Next, note that the computation of $\pbest$ and $\nbest$ (and hence which case
in $\caseset$ actually occurs) is not seen in the update
equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  This update is
usually implemented as a pair of independent if-statements, but that is a bit
awkward for our purposes.  We will instead make this mathematically explicit by
introducing a set of indicator functions, each representing a case from
Table~\ref{tab:evals}.  We can then sum over all of the cases, and the
indicator function will make all of the terms drop to zero except for the case
that actually occurs.

In what follows we will examine the indicator function and the position and
velocity update equations for a specific case, $\casexn$, as presenting
equations for every case would be tedious.  The extension to cases not
presented is straightforward.  When we are performing function minimization, we
can define the indicator function representing the case $\casexn$ by:

>>>>

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) techniques commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once, then
presents a comparison of the speculative algorithm with the original PSO.  In
\sec{relax}, we discuss various methods of improving the performance of
speculative evaluation in PSO, all of which break the requirement of strictly
reproducing the behavior of the original algorithm.  In \sec{conclusion} and
\sec{future} we conclude and discuss future work.

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \pURand_{t} \) and \( \nURand_{t} \) are vectors of independent random
numbers drawn from a standard uniform distribution, the \( \otimes \) operator
is an element-wise vector multiplication, $\pbest$ (called personal best) is
the best position the current particle has seen, and $\nbest$ (called
neighborhood best) is the best position the neighbors of the current particle
have seen~\cite{bratton-sis07}.  The parameters \( \ncoeff \), \( \pcoeff \),
and \( \constriction \) are given prescribed values required to ensure
convergence (2.05, 2.05, and .73, respectively)~\cite{clerc-tec02}.

Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies a particle is also a
neighbor to itself in that its own position and value are considered when
updating the particle's neighborhood best, $\nbest$.  Thus with $p$ particles,
using the Ring topology each particle with index $i$ has three neighbors:
$i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each particle has
$p$ neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in
\fig{fig:evals-sphere}, for some functions, and for large numbers of
processors, just adding particles reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
evaluation approach that allows us to perform iterations two at a time.

To do so we must first refactor PSO such that the determination of the value
of the objective function is separate from the rest of the computation.  For
simplicity, this discussion will describe the case where PSO is performing
function minimization using the Ring topology.  In this example, each particle
has two neighbors, the ``right neighbor'' and ``left neighbor,'' whose
positions are represented as $\rightn$ and $\leftn$ respectively.  Though we
will only describe the case of the Ring topology, this method can easily be
extended to arbitrary topologies.

The refactoring hinges on the idea that once the random coefficients are
determined, there are only a few possibilities for the new values of updates
to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible update
cases, identified in Table~\ref{tab:evals}.  We label each case with an
identifier referring to the source of the update: a minus sign ($-$)
represents no update, $L$ represents an update to $\nbest$ coming from the
left neighbor, $R$ represents an update to $\nbest$ coming from the right
neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$ coming
from the particle itself.  As an example, $\casexn$ refers to the case that
the particle finds a new personal best, but neither it nor its neighbors found
a position that updated its neighborhood best.  In the equations that follow,
we refer to an unspecified update case as $\casegen$, and to the set of cases
collectively as $\caseset$.

@@

For each case $\casegen \in \caseset$, we introduce an indicator function
$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
the cases, and the indicator function will make all of the terms drop to zero
except for the case that actually occurs.  Although this update is usually
implemented using if-statements, it is convenient to describe it as a
mathematical expression.  For example, the indicator function for the specific
case $\casexn$ can be written as follows:
